**Effective application of machine learning to microbiome based classification problems**

### 2019-10-07 submission/manuscript.Rmd
updates after Pats 2nd final read

---
title: '**Effective application of machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

\vspace{40mm}

@@

Prediction of health outcomes using machine learning (ML) is rapidly being adopted in human microbiome studies. However, these ML models are likely overoptimistic in terms of quantifying predictive performance. Moreover, there is a trend towards using black box models such as random forest and neural networks without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease. This work represents a step towards developing better ML practices in microbiome research by implementing a rigorous pipeline and emphasizing the importance of selecting ML models that reflect the goal of the study. These concepts are not particular to the study of health outcomes but can also be applied to environmental microbiology studies.

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases. However, the structure of the human microbiome is remarkably variable AMONG individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually. Recently, machine learning models (ML) have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017]. The task of diagnosing an individual relies on a rigorously validated model. However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress. These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a modelling approach has been due to an implicit assumption that more complex models are better. This has resulted in a trend towards using models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although complex models may in some cases, capture important non-linear relationships and therefore yield better predictions, they may result in black boxes that lack interpretibility. Such models require post hoc explanations to quantify the importance of each feature in making predictions. Depending on the goal of the modeling, researchers may choose to use different approaches. For example, researchers trying to identify the populations of microbiota associated wwith disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance. Nonetheless, it is important to understand the tradeoff and accepting that this tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018]. It is important for researchers to justify their choice of modelling approach.

@@

Model selection required tuning hyperparameters. Hyperparameters are parameters which need to be specified or tuned by the user, in order to train a model for a specific modeling problem. For example, when using regularization, (C) is a hyperparameter that indicates the penalty for overfitting. Hyperparameters are tuned using the training data to find the best model. We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0, where the model’s predictions are perfectly incorrect. An AUROC value of 0.5 indicates that model's predictions are no differect than random. To select hyperparameters, we performed a full grid search for hyperparameter settings when training our models. Default hyperparameter settings in previously developed ML packages in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization strengths (C).

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(format(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(format(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well non-linear models.

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that could indicate that the models were overfit to the training data. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. These differences are relatively small and gives us confidence in our estimate of the generalization performance of the models.

@@

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S3]. The variation in cross-validation performance within both models at lower sample sizes was larger than when the full collection of samples was used to train and validate the models. Because of the high dimensionality of the microbiome data (6920 OTUs), large sample sizes can lead to better models.

@@

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.  In this case, similar to previous studies [@sze_leveraging_2018; @wirbel_meta-analysis_2019], it was not possible to distinguish between health and disease using a single OTU. Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU was associated with classifying a subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S4]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. Although permutation analysis does not not allow us to determine the weight or the sign of the features, these results indicate that both methods are consistent in selecting the most important OTUs.

@@

To address these needs, we developed an open-sourced framework to ML models. using this pipeline, we benchmarked seven ML models and showed that the tradeoff between model complexity and performance may be less severe than originally hypothesized. In terms of predictive performance, the random forest model had the best AUROC compared to the other six models. However, the second-best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest is a non-linear ML model and is most often explained using post-hoc methods such as permutation importance. On the other hand, using L2-regularized logistic regression we ranked the importance of each OTU based on their feature weights. Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that predicts a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this data-split 100 times to measure the possible variation in predictive performance. During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the good test AUROCs.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among correlated features. It is important to know the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.

In this study, we did not consider all possible modeling approaches. However, the principles highlighted throughout this study apply to other ML modeling tasks with microbiome data. For example, we did not evaluate multicategory classification methods to predict non-binary outcomes. We could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories). We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs. Furthermore, as number of classes increases, more samples are required for each category to train an accurate model. We also did not use regression-based analyses to predict a non-categorical outcome. We have previously used such an approach to train random forest models to predict fecal short-chain fatty acid concentrations based on microbiome data [@sze_fecal_2019]. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Deep learning methods hold promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which can result in overfitting.

@@

**Data collection and study population.** The original stool samples described in our analysis were obtained from patients recruited by Great Lakes-New England Early Detection Research Network (Reference from Mack?). Stool samples were provided by adults who were undergoing a scheduled screening or surveillance colonoscopy. Participants were recruited from: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients’ colonic health was visually assessed by colonoscopy with bowel preparation and tissue histopathology of all resected lesions. We assigned patients into two classes: those with normal colons and those with screen relevant neoplasias (SRNs). The normal class included patients with normal colons or non-advanced adenomas whereas the SRN class included patients with advanced adenomas or carcinomas (reference for SRN?). Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas. There were 172 patients with normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the 198 adenomas, 109 were identified as advanced adenomas. Together 261 patients were classified as normal and 229 patients were classified as having a SRN.

@@

For hyperparameter selection, we started with a granular grid search. Then we narrowed and fine-tuned the range of each hyperparameter. A full grid search was needed to avoid large variation in prediction performance. For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has. For random forest, we tuned the __number of features__ to consider when looking for the best tree split. For XGBoost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners. Recently developed tools such as Hyperband help researchers with hyperparameter selection that can be incorporated to microbiome studies [@li_hyperband:_2016] .

@@

**Permutation importance workflow.** We calculated a Spearman's rank-order correlation matrix and defined correlated OTUs as having perfect correlation (correlation coefficient = 1 and p < 0.01). Non-correlated OTUs were permuted individually whereas correlated ones were grouped together and permuted at the same time.

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the AUROC values of the seven ML models by Wilcoxon rank sum tests to determine the best predictive performance.

@@

@@

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. ** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Predictive performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the median as the horizontal line in the box. The whiskers show the farthest points that were not outliers. Outliers were defined as those data points that are not within 1.5 times the interquartile ranges. 
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

@@

**Figure S5. Training times of ML models when dataset is subsetted .** (A) L2-regularized logistic regression and (B) Random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects. As the size of the dataset increased, the training times for L2-regularized logistic regression and random forest models increased linearly.

>>>>

---
title: '**Effective application of machine learning to microbiome based classification problems**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

\vspace{38mm}

@@

Prediction of health outcomes using machine learning (ML) is rapidly being adopted in microbiome studies. However, these ML models are likely overoptimistic in terms of quantifying predictive performance. Moreover, there is a trend towards using black box models such as random forest and neural networks without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease. This work represents a step towards developing better ML practices in microbiome research by implementing a rigorous pipeline and emphasizing the importance of selecting ML models that reflect the goal of the study. These concepts are not particular to the study of health outcomes but can also be applied to environmental microbiology studies.

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases. However, the structure of the human microbiome is remarkably variable among individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population seperately [@segata_metagenomic_2011]. Recently, machine learning models (ML) have grown in popularity among microbiome researchers because we are able to sample large numbers of individuals and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @mossotto_classification_2017]. The task of diagnosing an individual relies on a rigorously validated model. However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress. These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a modelling approach has been due to an implicit assumption that more complex models are better. This has resulted in a trend towards using models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although in some cases complex models may capture important non-linear relationships and therefore yield better predictions, they result in black boxes that lack interpretibility. Such models require post hoc explanations to quantify the importance of each feature in making predictions. Depending on the goal of the modeling, researchers may choose to use different approaches. For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance. Nonetheless, it is important to understand the tradeoff and accept that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018]. It is important for researchers to justify their choice of modelling approach.

@@

Model selection required tuning hyperparameters. Hyperparameters are parameters which need to be specified or tuned by the user, in order to train a model for a specific modeling problem. For example, when using regularization, C is a hyperparameter that indicates the penalty for overfitting. Hyperparameters are tuned using the training data to find the best model. We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2]. The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls. An AUROC value of 0.5 indicates that model's predictions are no differect than random. To select hyperparameters, we performed a full grid search for hyperparameter settings when training the models. Default hyperparameter settings in previously developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task. For example, of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization strengths (C).

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had higher test AUROC values than the other models for detecting SRNs (resampling test from two groups, p < 0.05). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(format(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(format(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models. To support our claim, we investigated the distribution of differences between random forest and L2-regularized logistic regression AUROC values for each data-split (Figure S3). Our results underscored that L2-regularized logistic regression perform nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that could indicate that the models were overfit to the training data. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. These differences are relatively small and give us confidence in our estimate of the generalization performance of the models.

@@

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S4]. The variation in cross-validation performance within both models at lower sample sizes was larger than when the full collection of samples was used to train and validate the models. Because of the high dimensionality of the microbiome data (6920 OTUs), large sample sizes can lead to better models.

@@

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.  In this case, similar to previous studies [@sze_leveraging_2018; @wirbel_meta-analysis_2019], it was not possible to distinguish between health and disease using a single OTU. Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU was associated with classifying a subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S5]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S5], 17 of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. Although permutation analysis does not not allow us to determine the weight or the sign of the features, these results indicate that both methods are consistent in selecting the most important OTUs.

@@

To address these needs, we developed an open-sourced framework to ML models. Using this pipeline, we benchmarked seven ML models and showed that the tradeoff between model complexity and performance may be less severe than originally hypothesized. In terms of predictive performance, the random forest model had the best AUROC compared to the other six models. However, the second-best model was L2-regularized logistic regression with a median AUROC difference less than 0.015 compared to random forest. While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest is a non-linear ML model and is most often explained using post hoc methods such as permutation importance. On the other hand, using L2-regularized logistic regression we ranked the importance of each OTU based on their feature weights. Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.

We established a pipeline that can be generalized to any modeling method that predicts a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this data-split 100 times to measure the possible variation in predictive performance. During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the good test AUROCs.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other, however, we could reduce the correlation threshold to further investigate the relationships among correlated features. It is important to know the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.

In this study, we did not consider all possible modeling approaches. However, the principles highlighted throughout this study apply to other ML modeling tasks with microbiome data. For example, we did not evaluate multicategory classification methods to predict non-binary outcomes. We could have trained models to differentiate between people with normal colons and those with adenomas or carcinomas (k=3 categories). We did not perform this analysis because the clinically relevant diagnosis grouping was between patients with normal colons and those with SRNs. Furthermore, as the number of classes increases, more samples are required for each category to train an accurate model. We also did not use regression-based analyses to predict a non-categorical outcome. We have previously used such an approach to train random forest models to predict fecal short-chain fatty acid concentrations based on microbiome data [@sze_fecal_2019]. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Deep learning methods hold promise [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018] but microbiome datasets often suffer from having many features and small sample sizes, which result in overfitting.

@@

**Data collection and study population.** The original stool samples described in our analysis were obtained from patients recruited by Great Lakes-New England Early Detection Research Network [@baxter_microbiota-based_2016]. Stool samples were provided by adults who were undergoing a scheduled screening or surveillance colonoscopy. Participants were recruited from: Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor (MI, USA). Patients’ colonic health was visually assessed by colonoscopy with bowel preparation and tissue histopathology of all resected lesions. We assigned patients into two classes: those with normal colons and those with screen relevant neoplasias (SRNs). The normal class included patients with normal colons or non-advanced adenomas whereas the SRN class included patients with advanced adenomas or carcinomas [@redwood_stool_2016]. Patients with an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology were classified as having advanced adenomas. There were 172 patients with normal colonoscopies, 198 with adenomas, and 120 with carcinomas. Of the 198 adenomas, 109 were identified as advanced adenomas. Together 261 patients were classified as normal and 229 patients were classified as having a SRN.

@@

For hyperparameter selection, we started with a granular grid search. Then we narrowed and fine-tuned the range of each hyperparameter. A full grid search was needed to avoid large variation in prediction performance. For L2-regularized logistic regression, L1 and L2-regularized SVM with linear and radial basis function kernels, we tuned the cost hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned sigma hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the depth of the tree where deeper the tree, the more splits it has. For random forest, we tuned the number of features to consider when looking for the best tree split. For XGBoost, we tuned for learning rate and the fraction of samples to be used for fitting the individual base learners. Recently developed tools such as Hyperband help researchers with hyperparameter selection that can be incorporated to microbiome studies [@li_hyperband:_2016] .

@@

**Permutation importance workflow.** We calculated a Spearman's rank-order correlation matrix and defined correlated OTUs as having perfect correlation (correlation coefficient = 1 and p < 0.01). OTUs without a perfect correlation to each other were permuted individually whereas correlated ones were grouped together and permuted at the same time.

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the AUROC values of the models by performing a resampling test. Our H0 model assumed that the AUROC values of two models were equal. To create the stocasthic model following H0, we permuted the AUROC values of the two models 10,000 times to be resampled to produce artificial datasets. We than compared the difference of means of the artificial resampled data and means of the real data.

@@

**Acknowlegements.**  We thank all the study participants of Great Lakes-New England Early Detection Research Network. We would like to thank the members of the Schloss lab for their valuable feedback during the formulation of the manuscript. Salary support for M.R came from NIH grant 1R01CA215574. Salary support for P.D.S. came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. ** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. The predictive performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the median as the horizontal line in the box. The whiskers show the farthest points that were not outliers. Outliers were defined as those data points that are not within 1.5 times the interquartile ranges. 
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

@@

**Figure S5. Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
